---
title: "Experiment 2"
author: "Owen Petchey"
date: "6/25/2021"
output: 
  html_document: 
    keep_md: yes
---


# Setup

## R

```{r}


rm(list = ls())

knitr::opts_knit$set(progress = TRUE, verbose = TRUE, cache = TRUE)

microxanox_release <- "0.2"

tmplib <- tempfile()
dir.create(tmplib)


### From '?remotes::install_github`:
# auth_token	
#   To install from a private repo, generate a personal access token (PAT) in
#   "https://github.com/settings/tokens" and supply to this argument. This is 
#   safer than using a password because you can easily delete a PAT without 
#   affecting any others. Defaults to the GITHUB_PAT environment variable.

remotes::install_github(
  "opetchey/microxanox",
  ref = microxanox_release,
  # auth_token = "ENTER YOUR TOKEN or PROVED AS ENVIRONMENT VARIABLE",
  build_vignettes = FALSE,
  force = TRUE,
  upgrade = FALSE,
  lib = tmplib
)

library(microxanox, lib.loc = tmplib)

library(tidyverse)
library(patchwork)
library(here)
source(here("experiments/r functions/various_useful_functions.r"))
zero <- 0 ## don't change
unity <- 1 ## don't change!!!
options(mc.cores = 8)
```

## Version of `microxinox` package used: `r packageVersion("microxanox") `

## General simulation conditions

```{r}
default_dynamic_model <- bushplus_dynamic_model
default_event_definition <- event_definition_1
default_event_interval <- 100
default_noise_sigma <- 0
default_minimum_abundances <- rep(1, 3)
names(default_minimum_abundances) <- c("CB", "PB", "SB") 
default_sim_duration <- 80000
default_sim_sample_interval <- 100
initial_pars_from <- "bush_ssfig3"
## note that next line (log10a_series is over-ridden with getting stable states)
#default_log10a_series <- c(-2, -2, -2, -2, -10, -10, -10, -10, -10)
```


## Define diversity

```{r}
num_CB_strains <- 9
num_SB_strains <- 9
num_PB_strains <- 9
CB_gmax_div <- 0.015789474
CB_h_div <- -0.08
SB_gmax_div <- 0.015789474 * 0.5
SB_h_div <- -0.323 * 0.5
PB_gmax_div <- 0.015789474 * 0.5
PB_h_div <- -0.323 * 0.5

num_div_treatment_levels <- 2
```

## Create diversity

```{r}
var_expt <- create_diversity()
```

## Display diversity

```{r}
display_diversity()
```


# Temporal switching

## Oxic to anoxic

### No diversity

```{r}
default_sim_duration <- 80000
```

```{r}
sim_number <- 1
default_log10a_series <- c(-1, -7, -7)
initial_state <- new_initial_state(num_CB_strains,
                                   num_PB_strains,
                                   num_SB_strains,
                                   values = "bush_ssfig3")
initial_state[grep("CB_", names(initial_state))] <- 10^10/num_CB_strains
sim_res_novar <- run_simulation(parameter_values = var_expt$pars[[sim_number]],
                          initial_state = initial_state)
plot_dynamics(sim_res_novar)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### Maximum diversity

```{r}
sim_number <- 2
sim_res_highvar <- run_simulation(parameter_values = var_expt$pars[[sim_number]],
                          initial_state = initial_state)
plot_dynamics(sim_res_highvar)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

```

## Anoxic to oxic

### No diversity

```{r}
default_sim_duration <- 60000
```


```{r}
sim_number <- 1
default_log10a_series <- c(-5, -3, -1, -1)
initial_state <- new_initial_state(num_CB_strains,
                                   num_PB_strains,
                                   num_SB_strains,
                                   values = "bush_ssfig3")
initial_state[grep("CB_", names(initial_state))] <- 10/num_CB_strains
sim_res_novar <- run_simulation(parameter_values = var_expt$pars[[sim_number]],
                          initial_state = initial_state)
plot_dynamics(sim_res_novar)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### Maximum diversity

```{r}
sim_number <- 2
sim_res_highvar <- run_simulation(parameter_values = var_expt$pars[[sim_number]],
                          initial_state = initial_state)
plot_dynamics(sim_res_highvar)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

```


## Anoxic to oxic to anoxic

### No diversity

```{r}
default_sim_duration <- 1000000
```

```{r}
default_minimum_abundances <- rep(100, 3)
names(default_minimum_abundances) <- c("CB", "PB", "SB") 
```

```{r}
sim_number1 <- 1
default_log10a_series <- c(-1, -8, -1)
initial_state <- new_initial_state(num_CB_strains,
                                   num_PB_strains,
                                   num_SB_strains,
                                   values = "bush_ssfig3")
initial_state[grep("CB_", names(initial_state))] <- 10/num_CB_strains
sim_res_novar <- run_simulation(parameter_values = var_expt$pars[[sim_number1]],
                          initial_state = initial_state)
plot_dynamics(sim_res_novar)
#ggsave(here("simulations/expt2/figures/switching_novar.pdf"), width = 10)
```

### Maximum diversity

```{r}
sim_number2 <- 2
sim_res_highvar <- run_simulation(parameter_values = var_expt$pars[[sim_number2]],
                          initial_state = initial_state)
plot_dynamics(sim_res_highvar)
#ggsave(here("simulationsexpt2/figures/switching_highvar.pdf"), width = 10)

```

### Visualise

```{r}
visualise_temporal_env_eco()
```



# Stable state finding

## Setup

```{r}
options(mc.cores = 6)
```


```{r}
ssfind_minimum_abundances <- rep(0, 3)
names(ssfind_minimum_abundances) <- c("CB", "PB", "SB")
ssfind_simulation_duration <- default_sim_duration
ssfind_simulation_sampling_interval <- ssfind_simulation_duration
ssfind_event_interval <- ssfind_simulation_duration
grid_num_a <- 1000 ## number of a_0 values
a_Os <- 10^seq(-7, -1, length=grid_num_a) ## sequence of a_0 values
grid_num_N <- 2 ## number of N values
initial_CBs <- 10^seq(0, 10, length=grid_num_N) ## sequence of N values
initial_PBs <- 1e8 ## not varied
initial_SBs <- 1e8 ## not varied
# next line creates all possible combinations
ss_expt <- expand.grid(N_CB = initial_CBs,
                      N_PB = initial_PBs,
                      N_SB = initial_SBs,
                      a_O = a_Os)


```

```{r}
#var_expt <- run_ss_var_experiment()
#saveRDS(var_expt, here("experiments/experiment 2/data/ss_data.RDS"))

```

```{r}
var_expt <- readRDS(here("experiments/experiment 2/data/ss_data.RDS"))


result_index1 <- 1
p1  <- plot_ss_result1(var_expt,
                result_index = result_index1,
                filename_prefix = NULL,
                save_image_file = FALSE)
p1
```

```{r}
result_index2 <- 2
p2  <- plot_ss_result1(var_expt,
                result_index = result_index2,
                filename_prefix = NULL,
                save_image_file = FALSE)
p2

```

```{r}
p_overlay <- plot_ss_result2(var_expt[result_index1,]$ss_res[[1]],
                             var_expt[result_index2,]$ss_res[[1]],
                             xlims = c(-7, -1))
p_overlay

ss_result1 <- var_expt[result_index1,]$ss_res[[1]]
ss_result2 <- var_expt[result_index2,]$ss_res[[1]]
xlims = c(-7, -1)
```

